---
title: "Typical AVNRT"
subtitle: Clinical Electrophysiology Conference
author: Anish S. Shah, MD, MS
institute: Division of Cardiology, University of Illinois at Chicago 
date: December 08, 2022
date-format: long
engine: knitr
format:
  revealjs:
    transition: fade
    transition-speed: fast
    slide-number: true
    incremental: false
revealjs-plugins:
  - attribution
jupyter: python3
---

```{r}
#| label: setup
#| echo: false
library(tidyverse)
library(coraline)
library(rhythm)
library(gganimate)
library(sf)
library(scales)
library(ggsci)

blue <- pal_material("blue", reverse = TRUE)(9)
orange <- pal_material("yellow", reverse = TRUE)(9)
red <- pal_material("pink", reverse = TRUE)(9)
green <- pal_material("teal", reverse = TRUE)(9)
grey <- pal_material("blue-grey", reverse = TRUE)(9)

find_project_folder <- function() {
	x <- sessionInfo()$running
	if (grepl("mac", x)) {
		file.path("/Users", "asshah4", "projects")
	} else if (grepl("Windows", x)) {
		file.path("C:/Users", "asshah4", "projects")
	}
}

loc <- file.path(find_project_folder(), "clinical", "lectures", "avnrt-series")
```

## Objectives

1. Understand the concepts of dual node physiology
1. Know the requirements for [typical AVNRT]{.orange}
1. Evaluate the response of [standard EP maneuvers]{.green} in [typical AVNRT]{.orange}

## {.smaller}

| Abbr. | Abbreviation |
|:--- |:---- |
| DAD | delayed after-depolarization |
| EAD | early after-depolarization | 
| APD | atrial premature depolarization | 
| VPD | ventricular premature depolarization | 
| RAAS | retrograde atrial activation sequence |

# Typical AVNRT

## {background-image="resources/josephsen-avnrt-schematic.png" background-size="contain"}

---

```{r}
#| label: normal with pathways
#| cache: true

# SF objects
chambers <- coraline::cardiac_atlas$right_atrium_rao$chambers
vessels <- coraline::cardiac_atlas$right_atrium_rao$vessels
projections <- coraline::cardiac_atlas$right_atrium_rao$projections
valves <- coraline::cardiac_atlas$right_atrium_rao$valves
septum <- coraline::cardiac_atlas$right_atrium_rao$septum
conduction <- coraline::cardiac_atlas$right_atrium_rao$conduction
reference <- coraline::cardiac_atlas$right_atrium_rao$reference
pathways <- coraline::cardiac_atlas$right_atrium_rao$pathways

# Basic figure
ra <- 
	st_union(chambers, projections, by_feature = TRUE) |>
	st_union(vessels, by_feature = TRUE) |>
	st_union(septum, by_feature = TRUE) |>
	st_union(valves, by_feature = TRUE) |>
	{\(.x) .x["geometry"]}() |>
	unique()

ra_grey <-
	ggplot(ra) +
	geom_sf(fill = grey[9], alpha = 0.1) +
	theme_void()

ra_grey +
	geom_sf(data = chambers, fill = red[5], alpha = 0.1) +
	geom_sf(data = septum, fill = red[5], alpha = 0.1) +
	geom_sf(data = vessels, fill = blue[5], alpha = 0.1) +
	geom_sf(data = projections, fill = red[2], alpha = 0.1) +
	geom_sf(data = valves, fill = grey[9], alpha = 0.1) +
	geom_sf(data = conduction, fill = orange[1], alpha = 0.3) +
	geom_sf(data = pathways, fill = "white", color = NA, alpha = 0.8) +
	geom_sf_label(data = filter(pathways, type == "pathway"), aes(label = label), nudge_y = 1)
```

---

```{r}
#| label: animated AVNRT
#| cache: true
avnrt <-
	ra_grey +
	geom_sf(data = pathways, fill = "white", color = NA_character_) +
	geom_sf(
		data = filter(pathways,
									structure == "slow" & component == "middle"),
		fill = orange[3]
	) +
	geom_sf(
		data = filter(
			pathways,
			(structure == "slow" & component == "distal") |
				(structure == "bridge" & component == "distal")
			),
		fill = orange[3]
	) +
	geom_sf(
		data = filter(pathways, structure == "lower_common"), 
		fill = orange[3]
	) +
	geom_sf(
		data = filter(pathways,
									structure == "fast" & component == "distal"),
		fill = red[2]
	) +
	geom_sf(
		data = filter(pathways,
									structure == "fast" & component == "middle"),
		fill = red[2]
	) +
	geom_sf(
		data = filter(
			pathways,
			(structure == "fast" & component == "proximal") |
				(structure == "bridge" & component == "proximal")
			),
		fill = red[2]
	) +
	geom_sf(
		data = filter(
			pathways,
			(structure == "upper_common" & component == "distal") |
				(structure == "upper_common" & component == "upper_fast")
			),
		fill = red[2]
	) +
	geom_sf(
		data = filter(pathways,
									structure == "upper_common" & component == "upper_slow"),
		fill = orange[2]
	) +
	geom_sf(
		data = filter(
			pathways, 
			(structure == "atrial_extension") |
				(structure == "slow" & component == "proximal")
		),
		fill = c(red[2], orange[3], rep(red[2], 4))
	) 

avnrt +
	transition_layers(
		keep_layers = c(Inf, Inf, rep(1, 9)),
		layer_length = 1,
		from_blank = FALSE
	) +
	enter_fade() +
	exit_fade(alpha = 0)
```

## Evaluation

1. Mode of initiation of tachycardia
1. Atrial activation sequence during tachycardia
1. Influence of bundle branch block on conduction and cycle length during tachycardia
1. Requirement of atria/ventricle for initiation and maintenance of tachcyardia
1. Effect of atrial/ventricular stimulation during tachycardia
1. Effect of drugs or physiological maneuvers on tachycardia

## Criteria for typical AVNRT {.smaller}

- Initiation and termination by APD/VPD during AV nodal Wenckebach cycles
- Dual AV nodal physiology (differential refractory curves during APD)
- Initiation dependent on critical AH interval during slow pathway conduction
- Retrograde atrial activation with variable VA activation in Triangle of Koch
- Initiated, terminated, or reset without atrial activation

::: aside
[Slow-fast AVNRT has atrial activation close to QRS complex, such that AH/HA > 1]{.grey}
:::

## {background-image="tmf_data/ecg-case-nsr.png" background-size="contain"}

## {background-image="tmf_data/ecg-case-svt-new.png" background-size="contain"}

---

::: {.columns}

::: {.column width="50%"}
__Sinus Rhythm__

- Normal P and R axis
- RR = 900
- PP = 900
- QRS = 80
- P = 120
- PR = 160
- QT = 440

:::

::: {.column width="50%"}
__Tachycardia__

- Suspected retro-grade P 
- RR = 350
- QRS = 70
- PR = 300  

::: {.fragment .highlight-red}
- RP = 40
:::

:::

:::

## Differential... {.smaller}

- Typical AVNRT
- Orthodromic AVRT
- Atypical AVNRT
- Atrial tachycardia _always a possibility_
- [Non-reentrant junctional tachycardia (junctional ectopic tachycardia)]{.grey}
- [Non-paroyxsmal junctional tachycardia (d/t digitalis-induced DADs)]{.grey}
- [Dual AV nodal tachycardia (repetitive retrograde concealment or "linking" phenomenon)]{.grey}

# Electrophysiology Study

## Case setup {.smaller}

::: {.columns}
::: {.column width="50%"}
Access: RCFV x 3

Catheters: 

- CS: decapolar catheter
- His: quadrapolar catheter
- RV: quadrapolar catheter
- Ablation: STSF 4 mm

Closure: collagen-plug x 3
:::

::: {.column width="50%"}
Study plan:

1. Baseline EP study
1. RA and RV pacing maneuvers (including decremental pacing)
1. His-refractory PVC
1. Induction of tachycardia
1. Entrainment
1. 3D anatomical mapping (CART) with His cloud
1. Ablation
1. Post-ablation testing
:::
:::

# Baseline

```{r}
#| label: LS pro data
#| include: false
nsr <- read_lspro(file.path(loc, "tmf_data/lsp-nsr-surface.txt")) 
tach_start <- 
	read_lspro(file.path(loc, "tmf_data/lsp-tach-initiation.txt")) 
cs_s1_640_dec <-
	read_lspro(file.path(loc, "tmf_data/lsp-cs-pace-decremental-640ms.txt")) 
cs_s1_400_jump <-
	read_lspro(file.path(loc, "tmf_data/lsp-cs-pace-decremental-400ms-jump.txt"))
abl_avn_erp <- 
	read_lspro(file.path(loc, "tmf_data/lsp-post-ablation-erp.txt")) 
avn_erp <- 
	read_lspro(file.path(loc, "tmf_data/lsp-avn-erp-s1-600ms-s2-220ms.txt"))
rv_burst_af <- 
	read_lspro(file.path(loc, "tmf_data/lsp-burst-into-af.txt"))
parahis <- 
	read_lspro(file.path(loc, "tmf_data/lsp-parahisian-pacing.txt")) 
tach_cycle <- 
	read_lspro(file.path(loc, "tmf_data/lsp-avnrt-ongoing.txt")) 
cs_entrain <- 
	read_lspro(file.path(loc, "tmf_data/lsp-cs-entrainment.txt")) 
rv_s1_500 <- 
	read_lspro(file.path(loc, "tmf_data/lsp-s1-500ms.txt")) 
rv_s1_300_entrain <- 
	read_lspro(file.path(loc, "tmf_data/lsp-entrainment-s1-300ms.txt")) 
his_pvc_fail <- 
	read_lspro(file.path(loc, "tmf_data/lsp-his-refractory-pvc-attempt.txt"))
his_pvc_win <- 
	read_lspro(file.path(loc, "tmf_data/lsp-his-refractory-pvc-success.txt"))
cs_s4_no_tach <- 
	read_lspro(file.path(loc, "tmf_data/lsp-s4-stim-no-tach.txt"))
```

---

```{r}
#| label: sinus rhythm from surface on EGM
#| cache: true
#| fig-cap: 12 lead sinus rhythm 
ggm(data = nsr,
		channels = c("I", "II", "III" , "AVR", "AVL", "AVF", paste0("V", 1:6)))
```

---

```{r}
#| label: normal intervals
#| cache: true
#| fig-cap: Baseline intervals showed AH of 82, and HV 50 at rate of 870 
ggm(data = his_pvc_win,
		channels = c("I", "CS", "RV", "HIS D", "HIS M"),
		time_frame = c(1.8, 2.2))
```

---

| Criteria | Status |
| :--- | --- |
| Initiation | - |
| Dual AV node | - |
| Critical AH during slow pathway | - |
| RAAS | [Retrograde P wave on surface]{.red} |
| Involvement of chambers | - |

# RA Pacing

---

```{r}
#| label: (post ablation) avn erp
#| cache: true
#| fig-cap: AV Wenckebach shown at 390
ggm(data = abl_avn_erp,
		channels = c("I", "CS", "HIS", "RV")) |>
	add_intervals(channel = "CS 9-10")
```

---

```{r}
#| label: avn erp
#| cache: true
#| fig-cap: AV node ERP at S1 = 600, S2 = 220
ggm(data = avn_erp,
		channels = c("I", "CS", "HIS", "RV")) |>
	add_intervals(channel = "CS 9-10")
```

---

```{r}
#| label: cs pacing 640
#| cache: true
#| fig-cap: Decremental pacing to evaluate dual AV nodal physiology
ggm(data = cs_s1_640_dec,
		channels = c("I", "CS", "HIS", "RV")) |>
	add_intervals(channel = "CS 7-8")
```

---

```{r}
#| label: cs pacing 400 with jump
#| cache: true
#| fig-cap: Demonstrated AH "jump" at around 500 ms (not shown).
ggm(data = cs_s1_400_jump,
		channels = c("I", "CS", "HIS", "RV")) |>
	add_intervals(channel = "CS 9-10")
```

---

| Criteria | Status |
| :--- | --- |
| Initiation | - |
| Dual AV node | [AH jump >50 with decremental pacing]{.red} |
| Critical AH during slow pathway | - |
| RAAS | [Retrograde P wave on surface]{.blue} |
| Involvement of chambers | - |

# RV Pacing

---

```{r}
#| label: Retrograde VA conduction
#| cache: true
#| fig-cap: RV shows retrograde conduction with variable VA timing
ggm(data = rv_burst_af,
		channels = c("I", "CS", "HIS", "RV"),
		time_frame = c(0.1, 1.5)) |>
	add_intervals(channel = "RV 1-2")
```

---

```{r}
#| label: RV pacing bursts into afib
#| cache: true
#| fig-cap: RV pacing lead to AF degeneration
ggm(data = rv_burst_af,
		channels = c("I", "CS", "HIS", "RV"),
		time_frame = c(1.2, 3.2)) |>
	add_intervals(channel = "RV 1-2")
```

---

```{r}
#| label: parahisian pacing
#| cache: true
#| fig-cap: Para-Hissian pacing at high output (when RBB selected, S-A interval widened)
ggm(data = parahis,
		channels = c("I", "CS", "HIS", "RV")) |>
	add_intervals(channel = "HIS D")
```

---

| Criteria | Status |
| :--- | --- |
| Initiation | - |
| Dual AV node | [AH jump >50 with decremental pacing]{.blue} |
| Critical AH during slow pathway | - |
| RAAS | [Concentric CS activation]{.red} |
| Involvement of chambers | - |

# Tachycardia Maneuvers

## Tachycardia initiation

---

```{r}
#| label: tach initiation
#| cache: true
#| fig-cap: Spontaneous initiation from APD
ggm(data = tach_start,
		channels = c("I", "CS", "RV", "HIS D", "HIS M"),
		time_frame = c(0.5, 3.5))
```

---

```{r}
#| label: tach zoom to APD
#| cache: true
#| fig-cap: Increased sweep speed of APD, suspected "bump" from His catheter
ggm(data = tach_start,
		channels = c("I", "CS", "RV", "HIS D", "HIS M"),
		time_frame = c(1.6, 2.9))
```

---

```{r}
#| label: continuous tach prior to pacing
#| cache: true
#| fig-cap: TCL of 375
ggm(data = tach_cycle,
		channels = c("I", "CS", "HIS", "RV"),
		time_frame = c(0.1, 2.5)) |>
	add_intervals(channel = "I")
```

## Tachycardia features prior to diagnostic pacing maneuvers

For typical AVNRT, there are some findings to help eliminate options.

1. VA Relationship: any pattern, but most commonly V=A and V<A
1. VA Interval: VA &le; 70
1. Atrial Activation Sequence: concentric
1. Spontaneous Termination: end either in A or V response
1. VA Increase with BBB: no

---

| Criteria | Status |
| :--- | --- |
| Initiation | [APD leads to tach]{.red} |
| Dual AV node | [AH jump >50 with decremental pacing]{.blue} |
| Critical AH during slow pathway | [AH interval is prolonging prior to APD]{.red} |
| RAAS | [Concentric CS activation]{.blue} |
| Involvement of chambers | - |

## His refractory VPDs

---

```{r}
#| label: his refractory pvc fail
#| cache: true
#| fig-cap: PVC placed between HV interval
ggm(data = his_pvc_fail,
		channels = c("I", "CS", "HIS", "RV"))
```

---

```{r}
#| label: his refractory pvc fail zoom
#| cache: true
#| fig-cap: PVC however affected the HH interval, thus not refractory
ggm(data = his_pvc_fail,
		channels = c("I", "CS", "HIS", "RV"),
		time_frame = c(1.2, 2.2))
```

---

```{r}
#| label: his refractory pvc success
#| cache: true
#| fig-cap: PVC breaks tachcyardia
ggm(data = his_pvc_win,
		channels = c("I", "CS", "HIS", "RV"))
```

---

```{r}
#| label: his refractory pvc success zoom
#| cache: true
#| fig-cap: Right bundle potential appears to be consistent in RV lead but direction is reversed after PVC
ggm(data = his_pvc_win,
		channels = c("I", "CS", "HIS", "RV"),
		time_frame = c(0.2, 2.0))
```

---

| Criteria | Status |
| :--- | --- |
| Initiation | [APD leads to tach]{.blue} |
| Dual AV node | [AH jump >50 with decremental pacing]{.blue} |
| Critical AH during slow pathway | [AH interval is prolonging prior to APD]{.red} |
| RAAS | [Concentric CS activation]{.blue} |
| Involvement of chambers | [No AP noted, VPD breaks cycle]{.red} |

## Entrainment

---

```{r}
#| label: cs entrainment
#| cache: true
#| fig-cap: Entrainment at 340 from CS ostia
ggm(data = cs_entrain,
		channels = c("I", "CS", "HIS M", "HIS D", "RV"),
		time_frame = c(0.5, 3.0)) |>
	add_intervals(channel = "CS 9-10")
```

---

```{r}
#| label: rv s1 entrained at attempted at 500
#| cache: true
#| fig-cap: Ventricular overdrive pacing (500) attempted but broke tachycardia
ggm(data = rv_s1_500,
		channels = c("I", "CS", "HIS M", "HIS D", "RV"))
```

---

```{r}
#| label: cs s1 entrained at 300
#| cache: true
#| fig-cap: Ventricular overdriving pacing at 300, VAHV response limited by VPD
ggm(data = rv_s1_300_entrain,
		channels = c("I", "CS", "HIS", "RV"))
```

---

| Criteria | Status |
| :--- | --- |
| Initiation | [APD leads to tach]{.blue} |
| Dual AV node | [AH jump >50 with decremental pacing]{.blue} |
| Critical AH during slow pathway | [AH interval is prolonging prior to APD]{.red} |
| RAAS | [Concentric CS activation]{.blue} |
| Involvement of chambers | [VPD breaks, A advances with H]{.red} |

# Conclusion

---

```{r}
#| label: post ablation s4
#| cache: true
#| fig-cap: After slow pathway modification, tachycardia non-inducible with isoproterenol (2 mcg) and multi-stim protocols.
ggm(data = cs_s4_no_tach,
		channels = c("I", "CS", "HIS", "RV"))
```

---

| Criteria | Status |
| :--- | --- |
| Initiation | [APD leads to tach]{.blue} |
| Dual AV node | [AH jump >50 with decremental pacing]{.blue} |
| Critical AH during slow pathway | [AH interval is prolonging prior to APD]{.red} |
| RAAS | [Concentric CS activation]{.blue} |
| Involvement of chambers | [VPD breaks, A advances with H]{.blue} |

Evidence supports typical AVNRT using slow fast pathway, without AP.
